---
title: "Basic objects"
output: 
  html_document: 
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Five basic or "atomic" classes of objects

* character
* numeric (real numbers - Double/Float)
* integer 
* complex (with imaginary parts)
* logical (TRUE/FALSE)

```{r}
a <- 2 # as numeric by default
typeof(a)
```

```{r}
a <- 2L # force integer
typeof(a)
```

```{r}
# special numbers 
a <- Inf # infinity
a <- NaN # undefined value - "not a number"
typeof(a)
```

### 1.1 `typeof` vs `class`

In R, `typeof()` and `class()` are both functions used to determine the type of an object, but they serve distinct purposes. 

`typeof()` function is used to determine the basic type (i.e. internal type and storage mode) of an R object (builtin objects). The possible values returned include: "character", "integer", "double", "logical", "raw" (for raw binary data), "complex", "list", "expression" (for expressions), "name", "symbol", "function", or "NULL" (for null/missing values). 

`class()` is associated with the object-oriented programming paradigm in R, where specific methods can be defined and dispatched based on the class of an object. It returns the class(es) or type(s) of an object, and provides information about the high-level, abstract category or defined class to which an object belongs. The return value of `class()` is a character vector containing the names of the classes or types an object inherits. 


## 2. Vectors

A vector can only contain objects of the same class. 
If objects in the `c()` function ("c" for concatenation) are not of the same class, then implicit coercion is applied. 

The direction of coercion is **logical > integer > double > character**, that is, if you combine logical and integer data types, the data type of vector will be integer, and so on. 

```{r}
x <- c("a", 2, T) # character vector
x <- c(3, FALSE) # numeric vector
```

### 2.1 Vector creation

Apart from using the `c()` function, vectors can be created using the `:` operator or the `seq()` function. In R programming, index starts from 1. 

```{r}
x <- c(1:8)
```

`seq()` function can be used to create more complex sequence, and can take the form of `seq(start, stop, step)` or `seq(start, stop, length.out=size)`. 

### 2.2 Explicit coercion using `as.*` functions, if available
```{r}
x <- 0:6
class(x)
as.numeric(x)
as.logical(x)
as.character(x)
```

Sometimes when R can't figure out how to coerce an object, `NA` can be resulted in with a warning from R. 
```{r}
x <- c("a", "b", "c")
as.numeric(x)
```

### 2.3 Matrices
Matrices are vectors with a dimension attribute. The dimension attribute is itself an integer vector of length 2 (number of rows, number of columns). Matrices cannot store different classes of objects. 

```{r}
m <- matrix(nrow = 2, ncol = 3)
m
dim(m)
attributes(m)
```

Matrices are constucted column-wise. Matrices can also be created directly from vectors by adding a dimension attribute. 
```{r}
m <- 1:10
dim(m) <- c(2, 5)
m
```

Matrices can be created by column-binding or row-binding with the `cbind()` and `rbind()` functions. 
```{r}
x <- 1:3
y <- 10:13
cbind(x, y)
```


## 3. Lists 
Lists are a special type of vector that can contain elements of different classes. Lists are a very important data type in R. They, in combination with the various "apply" functions, make for a powerful combination. 
```{r}
x <- list(1, "a", TRUE, 1 + 4i)
x
```

## 4. Factors
Factors are used to represent unordered or ordered categorical data. One can think of a factor as an integer vector where each integer has a label (corresponding to one level). Factors are important in statistical modeling and are treated specially by modeling functions like `lm()` and `glm()`. 

Often factors will be automatically created when you read a dataset in using a function like `read.table()`. These functions often default to creating factors when they encounter data that look like characters or strings. 

Factors can be created with the `factor()` function. 

```{r}
x <- factor(c("yes", "no", "yes", "no", "yes"))
x
table(x) # frequency table
## See the underlying representation of factor
unclass(x)
```

The order of levels of a factor can be set using the `levels` argument to `factor()`. This can be important in linear modeling because the first level is used as the baseline level. 

```{r}
x <- factor(c("yes", "no", "yes", "no", "yes"))
x ## levels in alphabetical order
## Assign user-defined order of levels 
x <- factor(c("yes", "no", "yes", "no", "yes"), levels=c("yes", "no"))
x
levels(x)
```

Factor variable trap: a factor is essentially an integer vector with levels. In R, it's easier to deal with numbers than character data. So R stores factors as integer vectors. The numbers are what R uses in the back end. But this creates a trap when you want to convert a character factor with each level being a number quoted as a string, to a numeric, as using `as.numeric()` will return the coded integers, not the original levels in numeric format. 

The correct way in this scenario is to convert the factor to character first, then convert to numeric. 

```{r}
a <- factor(c("20", "10", "30", "20", "30"))
typeof(a)
class(a)

b <- as.numeric(a)
b 

c <- as.numeric(as.character(a))
c

```

## 5. Data Frames
Data frames are represented as a special type of list where every element of the list has to be of the same length. So data frames can store different classes of objects in different columns. Each element of the list can be thought of as a column and the length of each element is the number of rows. 

Data frames are usually created by reading in a dataset using the `read.table()` or `read.csv()`. However, data frames can also be created explicitly with the `data.frame()` function, or they can be coerced from other types of objects like lists.  

```{r}
x <- data.frame(foo = 1:4, bar = c(T, T, F, T))
x
nrow(x)
ncol(x)
```

Data frames can be converted to a matrix by calling `data.matrix()`. 

```{r}
x <- data.frame(foo = 1:4, bar = c(T, T, F, T))
data.matrix(x)
```

Beaware that `data.frame()` default behavor is turn strings into factors.  Use `stringAsFactors = FALSE` to suppress this behavior. 

```{r}
df <- data.frame(x = 1:3, y = c("a", "b", "c"), stringsAsFactors = FALSE)
str(df)
```

## 6. Missing Values
Missing values are denoted by `NA` (or `NaN` for undefined mathematical operations). 

* `is.na()` is used to test objects if they are `NA`
* `is.nan()` is used to test for `NaN`. 
* `NA` also has a class, so there are integer `NA`, character `NA`, etc.
* `NaN` is also `NA`, but the converse is not true. 

```{r}
x <- c(1, 2, NA, NaN, 4)
is.na(x)
is.nan(x)
```

## 7. Names
R objects can have names, which is very useful for writing readable code and self-describing objects. 

### 7.1 Vectors with names
```{r}
x <- 1:3
names(x) <- c("New York", "a", "b")
names(x)
```

### 7.2 Lists with names
```{r}
x <- list("Los Angeles" = 1, Boston = 2, London = 3)
x 
names(x)
```

### 7.3 Matrices with row and column names
```{r}
m <- matrix(1:4, nrow = 2, ncol = 2)
dimnames(m) <- list(c("a", "b"), c("c", "d"))
m
```

Column names and row names can be set separately using the `colnames()` and `rownames()` functions. 

Comparison with data frames: 

| Object    | Set column names | Set row names |
| ------    | ---------------- | ------------- |
|data frame | `names()`        | `row.names()` |
| matrix    | `colnames()`     | `rownames()`  |








