---
title: "Subsetting R objects"
output: 
  html_document:
    toc: yes
---

There are three operators that can be used to extract subsets of R objects. 

* `[`: always return an object of the same class as original.  Can be used to select multiple elements of an object. 
* `[[`: used to extract a list or data frame. The class of the returned object will not necessarily be a list or data frame. Can only return a single element. 

* `$`: used to extract elements of a list or data frame by literal name. Its semantics are similar to that of `[[`.


## 1. Subsetting vectors 
Vectors can be subset using `[` operator. 
```{r}
x <- c("a", "b", "c", "d", "e")
x[1]

## extracting multiple elements
x[1:4]

## extracting multiple elements in arbitrary order
x[c(1, 3, 4)]

## passing a logical sequence to the `[` operator to extract elements meeting a given condition
x > "b"
x[x > "b"]
x[c(T, F, T, T)]
```

## 2. Subsetting matrices
```{r}
x <- matrix(1:6, 2, 3)
x

# extracting one element
x[1, 2]

# extracting a row
x[1, ]

# extracting a column
x[, 2]

```

#### 2.1 dropping dimensions
By default, when a single element (or row or column) of a matrix is retrieved, it is returned as a vector of length 1 rather than `1 x 1` matrix. Often this is exactly what we want, but this behavior can be turned off by setting `drop = FALSE`. 

```{r}
x <- matrix(1:6, 2, 3)
x[1, 2]
x[1, 2, drop = FALSE]

## extracting a single row
x[1, , drop = FALSE]
```

Be careful of R's automatic dropping of dimensions.  This is a feature that is often quite useful during interactive work, but can later come back to bite you when you are writing longer programs or functions. 

## 3. Subsetting lists (or data frames)
Lists can be subset using all three of the operators mentioned above, and all three are used for different purposes. 
The `[[` operator can be used to extract single elements from a list. You can also use the `$` operator to extract elements by name.  The difference between `[[` and `$` is that the `[[` operator can take computed indices, whereas the `$` operator can only be used with literal names. 

```{r}
x <- list(foo = 1:4, bar = list(0.6, "a", T))
x[[1]]

## by literal name
x[["bar"]]

## using $ operator 
x$foo # note that quotes are not needed when using the $ operator

## using computed indices
name <- "foo"
x[[name]]
```

#### Subsetting nested elements of a list

```{r}
x <- list(foo = 1:4, bar = list(0.6, "a", T))

## Get the 3rd element of the 2nd element
x[[c(2, 3)]]

## same as above
x[[2]][[3]]
```

#### Extract multiple elements of a list using `[` operator

```{r}
x <- list(foo = 1:3, bar = 1.3, baz = "New York")
x[c(1, 3)]
```

Remember that the `[` operator always returns an object of the same class. 

## 4. Remove NA values

A common task in data analysis is removing missing vlaues (NA's). 

```{r}
x <- c(2, 3, NA, 4, NA)
bad <- is.na(x)
x[!bad]
```

You can use the `complete.cases` function to remove elements or rows with any NA value. 

```{r}
x <- c(2, 3, NA, 4, 4, 5)
y <- c(NA, 2, 3, NA, NA, 3)
good <- complete.cases(x, y)
good
x[good]
y[good]
```

Using `complete.cases` on data frames:
```{r}
head(airquality)
good <- complete.cases(airquality) # returns a logical vector that aligns with the rows in the data frame
typeof(good) 
head(airquality[good, ])
```







