---
title: "Vectorized Operations"
output: 
  html_document:
    toc: yes
---


Many operations in R are vectorized, meaning that operations occur in parallel in certain R objects. This allows you to write code that is efficient, concise, and easier to read than in non-vectorized languages. Without vectorization, you would have to do a for-loop or something like that. 

```{r}
x <- 1:4
y <- 5:8
x + y
x >= 2
y == 6
x * y
x / y
```

## 1. Vector recycling

If two vectors are not of the same length in an operation, then the shorter vector will be recycled to match size of longer vector. 

```{r}
first <- c(1, 2)
second <- c(10, 20, 30, 40)
first + second
```

Delete a Vector in R: 
```{r}
j <- NULL
```

Sort a vector: 
```{r}
x <- c(1, 2, 22, 11, 33)
sort(x)
sort(x, decreasing = TRUE)
```


## 2. Vectorized matrix operations. 

Matrix operations can also be vectorized. This way we can do element-by-element operations on matrices without having to loop over every element. 

```{r}
x <- matrix(1:4, 2, 2)
y <- matrix(rep(10, 4), 2, 2)

## element-wise multiplication
x * y

## true matrix multiplication
x %*% y 
```


