
---
title: "Dates and Times in R"
output: html_document
---

R has developed a special representation for dates and times. Dates are represented by the `Date` class and times are represented by the `POSIXct` or `POSIXlt` class. Dates are stored internally as the number of days since 1970-01-01, while times are stores internally as the number of seconds since 1970-01-01. 

## 1. Dates in R

Dates are represented by the `Date` class and can be coerced from  a character string using the `as.Date()` function. This is a common way to end up with a `Date` object in R. 

```{r}
d <- as.Date("1971-01-01")
class(d)
typeof(d)

## view the internal representation of a Date object by using the unclass function
unclass(d)
```

## 2. Times in R

Times are represented by the `POSIXct` or `POSIXlt` class. 

* `POSIXct` is just a very large integer under the hood. It is a useful class when you want to store times in something like a data frame. 
* `POSIXlt` is a list underneath.  It stores a bunch of other useful information like the day of the week, day of the year, month, day of the month. It is useful when you need this kind of information.

Times can be coerced from a character string using the `as.POSIXct` or `as.POSIXlt` function. You can coerce back and forth using these two functions between `POSIXct` and `POSIXlt`. 

```{r}
x <- Sys.time() ## This returns a POSIXct format time
x 
typeof(x)
class(x) ## POSIXct
unclass(x)
names(unclass(x))

## coerce to POSIXlt as a list
p <- as.POSIXlt(x)
class(p)  ## POSIXlt
unclass(p)
names(unclass(p))
p$wday
```

You can use the `strptime()` function to convert dates and times written in a character string format into a times format using a format string. You can check the details of format string by `?strptime`. 

```{r}
datestring <- c("January 10, 2021 11:40", "December 8, 2020 3:40")
datestring
t <- strptime(datestring, "%B %d, %Y %H:%M")
t
class(t) # POSIXlt
```

## 3. Generic functions that work on dates and times

For example, there are the following generic functions. 

* `weekdays`: give the day fo the week
* `months`: give the month name
* `quarters`: give the quarter number ("Q1", "Q2", "Q3", or "Q4")

These generic functions work on both `Date` class and `POSIXct` or `POSIXlt` class. 


## 4. Operations in dates and times

You can use mathematical operations such as + and - on dates and times. Note that you can't perfom operations on `Date` and `POSIXlt`.  To do it, you will have to convert a `Date` to `POSIXlt` using the `as.POSIXlt()` function. 

You can also do comparisons too (i.e. ==, <=, >=).

The nice thing about th date/time classes is that they keep track of all the annoying things about dates and times, like leap years, leap seconds, daylight savings, and time zones. 

```{r}
x <- as.POSIXct("2021-12-11 01:00:00")
y <- as.POSIXct("2021-12-11 07:00:00", tz = "GMT")
y-x
```
