
---
title: "Control Structures in R"
output: html_document
---

Commonly used control structures in R are

* `if` and `else`: testing a condition and act on it
* `for`: execute a loop a fixed number of times
* `while`: execute a loop while a condition is true
* `repeat`: execute an infinite loop (must `break` out of it to stop)
* `break`: break the execution of a loop
* `next`: skip an interation of a loop

## 1. `if-else`

```
if (<condition>) {
    ## do something
} else if  (<condition2>) {
    ## do something different
} else {
    ## do something different
}
```

Of course, the `else` claus is not necessary. 

Sometimes there is more than one condition. Conditions are always evaluated from left to right. In the example below, if `x` were less than 3, the condition `x < 11` would not have been evaluated. 

```
if (x >=3 && x < 11) {
    ## do something
}
```

## 2. `for` loops

For loops are pretty the only looping construct that you will need in R. 

In R, for loops take an iterator variable and assign it successive values from a sequence or vector. For loops are most commonly used for iterating over the elements of an object (list, vector, etc.). 

#### 2.1 
```
x <- c("a", "b", "c", "d")

for (i in 1:4) {
    print(x[i])
}
```

#### 2.2 

You can use `seq_along()` and `seq_len()` in conjunction with for loops.  These two functions return an integer vector. (Unless it's a long vector when it will be double.)

```
for (i in seq_along(x)) {
  print(x[i])
}
```

#### 2.3 

It is not necessary to use an index-type varaible. 

```
for (letter in x) {
    print(letter)
}
```

#### 2.4 Nested for loops

```{r}
x <- matrix(1:4, 2, 2)

for (i in seq_len(nrow(x))) {
    for (j in seq_len(ncol(x))) {
        print(x[i, j])
    }
}
```

The above code is equivalent to the following: 

```{r}
x <- matrix(1:4, 2, 2)

for (e in x) {
  print(e)
}
```

Nested loops are commonly needed for multidimensional or hierarchical data structures (e.g. lists, matrices). However, nesting over 2 to 3 levels often makes it difficult to read/understand the code. 

If you find yourself in need of a large number of nested loops, you may want to break up the loops by using functions. 


## 3. `while` loops

While loops can potentially result in infinite loops if not written properly. Use with care!

Example: 

```
count <- 0 
while (count < 10) {
    print(count)
    count = count + 1
}
```

## 4. `repeat` loops

`repeat` initiates an infinite loop right from the start. These are not commonly used in statistical or data analysis applications but they do have their uses. The only way to exit a `repeat` loop is to call `break`. 

For example, 

```
x0 <- 1
tol <- 1e-8

repeat {
    x1 <- computeEstimate() ## made-up function
    if (abs(x1 - x0) < tol) {
        break
    } else {
        x0 <- x1
    }
}
```

The loop above is a bit dangerous as there's no guarantee it will stop. You could get in a situation where the values of `x0` and `x1` oscillate back and forth and never converge. Better to set a hard limit on the number of iterations by using a `for` loop and then report whether convergence was achieved or not. 


## 5. `next`, `break`

`next` is used to skip an iteration of a loop. 

`break` is used to exit a loop immediately, regardless of what iteration the loop may be on. 














